Scalable analysis of genetic variation data in Python

### 2024-01-10 paper.tex
Update on storage section

\begin{figure}
    \begin{tabular}{c}
% FIXME temporarily copying this in here until Tim's tikz code arrives.
% https://github.com/pystatgen/sgkit-publication/issues/46
\includegraphics[width=7cm]{diagrams/tmp-illustration.png} \\
\includegraphics[width=7cm]{figures/data-scaling}
\end{tabular}
\caption{Chunked compressed storage of the genotype matrix using Zarr.
(A) Schematic of the chunked array storage used in sgkit
[Note: wrong fig. See
\url{https://github.com/pystatgen/sgkit-publication/issues/46} for
proposed update].
(B) Compression performance compared with BCF, savvy and genozip
based on simulated data (see text for details).
\label{fig-data-storage}}
\end{figure}

@@

\subsection{Storing genetic variation data}
The shortcomings of VCF (and to a lesser degree, BCF) as a storage
format and the basis of efficient computation have long been
recognised \citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many alternatives
have been proposed.

>>>>

\subsection{Storing genetic variation data}
Storage is the foundation on which any large-scale data-driven computation
rests: how we encode and organise data on persistent storage (the format)
fundamentally affects and limits the efficiencies of downstream calculations.
Different formats are optimised for different purposes, and there are 
often many competing requirements when designing a data storage format.
% TODO is there some literature we could cite on this? I'm making up the 
% term "computational format" here, presumably there is lots  prior thought
% on these ideas?
For our  purposes here, it is important to distinguish between \emph{archival} 
and \emph{computational} formats. An archival format is optimised for 
maximum interoperability and future-proofing, and uses the simplest and 
most robust technologies to ensure the long-term usability of the data.
Reasonable compression levels are important because 
of storage and transfer costs, but the efficiency of data access is a 
secondary concern. 
A computational format, on the other hand, is optimised 
for calculations, for \emph{using} the data, today. A computational format 
is designed with particular data access patterns in mind, ultimately
to make a broad range of queries and downstream computations as efficient as 
possible.

% Note: a quick explanation of VCF for CS people coming in fresh
VCF~\citep{danecek2011variant} is a text-based format in which each line 
describes observations 
and metadata for a single variant (a position on the genome at which
variation occurs in the sample). The first eight tab-delimited 
columns are mandatory, and subsequent columns describe the per-sample 
genotype calls and quality control data.

[WIP: currently working on weaving together the stuff below into
some sort of narrative]

As has been noted many times~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},

The shortcomings of VCF (and to a lesser degree, BCF) as a storage
format and the basis of efficient computation have long been
recognised 
and many alternatives
have been proposed.

\begin{figure}
    \begin{tabular}{c}
% FIXME temporarily copying this in here until Tim's tikz code arrives.
% https://github.com/pystatgen/sgkit-publication/issues/46
\includegraphics[width=7cm]{diagrams/tmp-illustration.png} \\
\includegraphics[width=7cm]{figures/data-scaling}
\end{tabular}
\caption{Chunked compressed storage of the genotype matrix using Zarr.
(A) Schematic of the chunked array storage used in sgkit
(B) Compression performance compared with BCF, savvy and genozip
based on simulated data (see text for details).
\label{fig-data-storage}}
\end{figure}

@@
